---
title: "Periop Wearables: Heart Rate Data from FitBit"
output: github_document
author: "Dr Chris Tomlinson"
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning=FALSE, message=FALSE)
```

In Part 1 we explored using the [fitbitr](https://github.com/teramonagi/fitbitr) package from [Nagi Teramo(teramonagi)](https://github.com/teramonagi) and the [fitbit Web API](https://dev.fitbit.com/build/reference/web-api/) to extract **Heart Rate Time Series** Data.

In Part 2 we will use a *Personal* application API key to access Intraday Time Series Data, giving us heart rate data with a greater temporal resolution e.g. detail-level = 1sec OR 1min.
```{r loadtemptoken, echo=FALSE}
# Uncomment the line below and run in console to create the temporary token
#saveRDS(token, "temp_token")
token <- readRDS("temp_token")
date <- "2020-05-25"
# Also need to load fitbitr again as the above chunk was not evaluated
library(fitbitr)
```
## Intra-day Time Series

We will use the `get_heart_rate_intraday_time_series()` function which has the following arguments:  
* `date` The date, in the format `yyyy-MM-dd` or `today`.  
* `detail_level` Number of data points to include. Either `1sec` or `1min`. Optional.  
* `start_time` The start of the period, in the format `HH:mm`. Optional.  
* `end_time` The end of the period, in the format `HH:mm`. Optional.  
 
Let's look at the structure of the data this function returns:
```{r getdata}
library(fitbitr)
str(get_heart_rate_intraday_time_series(token, date = date, detail_level = "1min"))
```
Pretty straightforward, we can see a `data.frame` consisting of 2 variables; a character vector `time` (from `start_time` to `end_time`, defaulting to a 24-hour period from "00:00:00" on the `date` specified) and an integer vector `value` corresponding to the Heart Rate for that time.

Let's import the data into a data.frame and rename `value` to the more descriptive `HR`. We must also convert `time` from a character to something more correct. Otherwise when we call `geom_smooth()` in `ggplot` it won't produce any result. I've used the `as.difftime()` function in base R, rather than the more common `strptime()`, as it gives me a time since midnight in units of my choice, rather than appending the time to todays date which would introduce confusion when analysing data from multiple days.

```{r cleandata}
df <- get_heart_rate_intraday_time_series(token, date = date, detail_level = "1min")
# Rename col 'value' to 'HR
names(df)[2] <- "HR"
# Convert time from class = character to difftime, i.e. mins since 00:00
df$time <- as.difftime(df$time, units = "mins")

```

## Plotting Intra-day Time Series  

We can plot this data as a simple scatter plot in `ggplot` using `geom_point()`. I've coloured it red, given the cardiac nature of the data, and changed it's transparency `alpha = 0.4` to improve legibility given the close distribution of the points along the x axis.

Finally I've added a smoother (? correct term) using `geom_smooth()` to aid the eye in seeing patterns given the 'overplotting' described above. This defaults to `method = "gam"` with `formula = y ~ s(x, bs = "cs")`. I don't have the depth of mathematical knowledge to say if this is a correct method, but visually 'eyeballing' the resultant line it looks pretty appropriate to me!

I've elected to show a 95% (default) Confidence Interval using the argument `se = TRUE` . `labs()` will add a title and label the x-axis, helping to clarify that time is in minutes as `difftime` is not the most obvious.
```{r hrplot}
library(ggplot2)
library(ggthemes)
p <- ggplot(df, aes(x = time, y = HR)) + geom_point(col = "red", alpha = 0.4) + geom_smooth(se = TRUE) + labs(title = "Heart Rate over Time", x = "Time (mins)") + theme_few()
p
```

## Exploring different temporal resolutions

As described above the `get_heart_rate_intraday_time_series()` function has an argument `detail_level` which specifies the 'Number of data points to include', i.e. the temporal resolution. This can be specified as either `1sec` (and seemingly not any other number of minutes) or `1min`.

If we could reduce the temporal resolution even from 1 minute to 2 minutes this would halve the amount of data to process, particularly important when we start analysing larger time periods and more than one subject! Clearly this will come at the expense of a loss of detail so I thought it would be interesting to plot the same heart rate data at different detail levels (1 min, 5 min & 15 min) and compare the results. Particularly the smoothed line.

Unfortunately `get_heart_rate_intraday_time_series()` won't accept a `vector` for `detail_level` so we have to resort to separate function calls, storing the outputs in their own data frames. We will then `merge()` these (again unfortunately only can merge two at a time, so have to repeat twice) into a single dataframe, using the arguments `all = TRUE` as we don't want to select only those time points that intersect (this would lose the greater resolution, defeating the point of the exercise!). I had specified `no.dups = FALSE` although in retrospect this is unecessary as they're all pulling data from the same source so the temporal resolution shouldn't change the heart rate.

We then need to do a bit of tidying, we will rename the columns to their `detail_level`: `"HR_1min", "HR_5min", "HR_15min"` respectively and convert the `time` to `difftime` to allow us to perform the `geom_smooth()` regression line.

```{r detaillevels}
df.1m <- get_heart_rate_intraday_time_series(token, date = date, detail_level = "1min")
df.5m <- get_heart_rate_intraday_time_series(token, date = date, detail_level = "5min")
df.15m <- get_heart_rate_intraday_time_series(token, date = date, detail_level = "15min")

df.detail.levels <- merge(df.1m, df.5m, by = "time", all = TRUE, no.dups = FALSE)
df.detail.levels <- merge(df.detail.levels, df.15m, by = "time", all = TRUE, no.dups = FALSE)

names(df.detail.levels)[2:4] <- c("HR_1min", "HR_5min", "HR_15min")
df.detail.levels$time <- as.difftime(df.detail.levels$time, units = "mins")
```

There would be various ways of representing the resultant information. I could produce a 1x3 panel plot. Instead I decided to follow the principle of showing as much useful info in a single figure as possible and instead plot a line for each `detail_level` on the same axes.

The data is currently in 'Wide Format':
```{r}
head(df.detail.levels, 3)
```
It is easier for us to plot if we convert it into 'Tall Format' using the `melt()` function in the `reshape2` package with the argument `id.vars = "time"` to show that we are preserving time for our x-axis.
```{r melt}
library(reshape2)
melted <- melt(df.detail.levels, id.vars = "time")
head(melted, 3)
```
Now we can create the plot in `ggplot` using the aesthetic `col = variable` to colour the points by the melted column `variable` which consists of the `detail_level`.
```{r 1second}
ggplot(melted, aes(x = time, y = value, col = variable)) + geom_point(alpha = 0.5) + geom_smooth(aes(col = variable), se = FALSE) + labs(title = "Heart Rate over Time", subtitle = "By varying detail level", x = "Time (mins)", y = "HR", colour = "Detail Level") + theme_few()
```
  
## Discussion  
We can see the following observations:  
  
* A greater detail level gives a greater range of values. This is most significant for the higher heart rates, presumably because they are more transient, in contrast to the low heart rates which were probably sustained during sleep.  
* Overall there is little difference in the averaged data between the various detail levels.  
 
This is all fairly intuitive and we can also be seen using the extremely useful `summary()`
```{r}
summary(df.detail.levels)
```
  
## Conclusions  
>I would thus conclude that capturing heart rate data at 15-minute intervals does not result in a significant loss of detail, **provided we are not particularly interested in looking at the peak heart rates**.